%!TEX root = ./main.tex
\section{Related Work}\label{sec:related}

% \paragraph{\gql.} 
To the best of our knowledge, the only formalization efforts around \gql are \HP~\cite{gqlph} and~\cite{olafschema}, which we have already discussed. The rest of the \gql literature focuses on practical issues such as creating \gql services and migrating REST-based web services \gql~\cite{improvingoeeu, ehriapi, gqlexperiences}, automatic migration~\cite{migratingapi}, and testing techniques~\cite{gqldeviation}. A couple of empirical studies analyze the structure of \gql schemas in commercial and open-source projects~\cite{empiricalgql, empiricalapi}, shedding interesting insights on \gql in practice.
% Although these provide , they do not provide guarantees \wrt the specification's correctness. 
% As far as we are aware, the only works on formalizing \gql are the present work and~\cite{gqlph}.
% Most of the academic research on \gql 
% Others focus on testing techniques for implemented \gql services or tools for 

% \paragraph{Mechanization of query languages.} 
The formalization of data management systems has received a lot of attention in the traditional 
relational data model~\cite{relationalcoq}, SQL and its semantics~\cite{sqlequiv, hottsql, vesqlengines, vesqlsemantics}, as well as related query languages and engines~\cite{certifdatalog}. The tree-based nature of \gql queries and response differs significantly from the tuple-based semantics in traditional query languages, requiring different models and techniques.
% Although these efforts extensively cover these models and languages, we believe they are 
% not directly applicable to the \gql setting\td{One of the reasons is that queries and responses in \gql are both trees (the atomic response IS a tree), while }. 
% \td{Idk what to say tbh. I believe most of this work doesn't directly apply to our context, but it exists and cover several topics.}
% \td{Also, I am not an expert in databases and query languages, so I am not entirely sure how or what to compare}
% \td{I remember having a talk with Jorge where he mentioned that \gql is a lot simpler and that, although he thought they can be related to AQL, 
% they might not, because of this tree form of queries and responses -- which differs completely from the list of tuples in other query languages}
% \paragraph{Mechanization of graph data model.} 
Doczkal and Pous~\cite{graphtheory} develop a mechanization of graph theory in \coq, including simple graphs, digraphs, and their properties. Their work could possibly be extended to deal with property graphs, and used for \gcoql.
Bonifati \etal~\cite{graphviewmaint} study graph view maintenance. Their definition of graphs is tailored to handle regular Datalog queries and does not entirely fit the \gql setting; it could however serve as a base to extend \gcoql with mutation.

% There is also work on 
% \td{They define a function that given a regexp returns a whole graph of the nodes that are connected via the regexp. Nodes are not typed, nor do they have properties I believe (not sure).}

% \paragraph{\gql miscellaneous.} 
Finally, despite being used mostly for web services, there are efforts to extend notions used in \gql to other areas of database specification and querying. Hartig and Hidders~\cite{olafschema} use the \gql schema definition DSL to define the structure of property graphs, which can be linked to similar efforts to define the structure of graph databases~\cite{schemaval}. Taelman \etal study the transformation of \gql queries to SPARQL~\cite{gqlsparql}.\et{is there any mechanized formalization of sparql?}\td{Haven't found one yet!}
% using \gql for deductive database in a Prolog setting~\cite{gqldeductive}. 
% \td{Not entirely sure what the last one is tbh, so probably	kill xd}


\td{Mention the JSON formalization done by reutter?~\cite{json}}\et{what could we do with it? is there anything interesting to say? (eg. for responses?)}
\td{Not sure. I looked at it when I was confused with HP's responses and thought I could base it off of it, but their formalization is a bit different than what I was expecting.
They mention that theirs is the first formalization, but idk}

